---
title: "Principal Component Analysis (PCA)"
output: html_document
---

**Suppose the matrix X is already given in R, then you can compute the scatter matrix (S) simply by:**

$$S = X'X$$

```{r}
X <- matrix(c(1,-3,1,-3,-1,2,1,2,3), ncol=3, byrow = TRUE)
X
```

```{r}
S <- t(X) %*% X
S
```

**If you want to compute the eigenvectors and eigenvalues of the matrix**

```{r}
M <- matrix(c(1,-3,1,-3,-1,2,1,2,3) , ncol=3, byrow = TRUE)
M
```

```{r}
eigen_of_M <- eigen(M)
eigen_of_M
```

**Now, we enter the dataset into R, and use prcomp() to carry out PCA.**

- We first do centering but not normalization.

```{r}
Y <- matrix(c(-3, 2.2, 4, 1.1, -5.8, -2, 1.9, 3, 1.2, -4.1,
               0, 2.1, 1, 1.2, 0, 1, 1.7, 0, 1.2, 2.1,
               2, 1.8, -1, 1.2, 3.7, 4, 2.3, -3, 0, 7.9,
               6, 2.2, -5, 1.2, 12.3), ncol=5, byrow = TRUE)
Y
```

```{r}
pca.noscale <- prcomp(Y,center = TRUE)
pca.noscale
```

**Next, we do both centering and normalization.**

- scale = TRUE => do centering and do normalization

```{r}
pca.scale <- prcomp(Y, scale = TRUE)
pca.scale
```

#### Quiz

For this question, you may use R or Python. Your answer must be accurate to at least 4 decimal places.

What is the sum of all eigenvalues of 

```{r}
q <- matrix(c(13.3, -6.4, 1.8, -17.6, -6.4, -7.7, 1.8, -10.1, 1.8,
              1.8, -4.4, 27.9, -17.6, 10.1, 27.9, 5.2), 
            ncol=4, byrow = TRUE)
q
```


```{r}
eigen_of_q <- eigen(q)
eigen_of_q
```


```{r}
eigen_of_q$values
35.333241+-30.169024 + 10.375158 + -9.139374
sum(eigen_of_q$values)
```


----

----